Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-08-03 manuscript.tex
results section edits

\subsection*{Evolutionary relationships of sex chromosome races in \textit{R. hastatulus}}

@@

%figure

%Figure 4. Evolutionary relationships of sex chromosome races in Rumex hastatulus, inferred using the Neighbor-Joining method (Saitou and Nei 1987). The percentage of replicate trees in which the associated sequences clustered together in the bootstrap test (1000 replicates) is shown next to branches (Felsenstein 1985). The phylogeny is drawn to scale, with branch lengths in the same units as those of the evolutionary distances used to infer the tree. The evolutionary distances were computed using the Maximum Composite Likelihood method (Tamura et al. 2004) and are in the units of the number of base substitutions per site. The analysis was conducted on an alignment of X- and Y-linked genes from Rumex hastatulus, with orthologous autosomal sequences from the non-dioecious but closely related outgroup species Rumex bucephalophorus used to root the tree. ‘N’ designates the North Carolina race, and ‘T’ the Texas race, with numbers 1-6 corresponding to each of the 6 populations sampled. The inferred SC (South Carolina), and FL (Florida) sub-clades are indicated.

>>>>

\subsection*{Y-chromosome diversity in \textit{R. hastatulus} is very low}
Our analysis of polymorphism in \textit{R. hastatulus} revealed that diversity on the Y chromosome is significantly lower than neutral predictions, with estimates indicating $Y/A=0.02$, a 12.5 fold reduction compared to the neutrally expected $Y/A = 0.25 (P<0.0001)$. This corresponds to a 50-fold reduction compared to the mean diversity estimated on autosomes (\X), and a 40-fold reduction compared to X chromosome diversity (Table 1). By normalizing X- and Y-chromosome diversity estimates by autosomal diversity, these results suggest that the X-Y difference we observe (an \X fold difference) is not due to some process specifically elevating diversity on the X chromosome. In addition, a recent analysis of sex-linked substitution rates in this species found that the number of synonymous mutations in X and Y lineages, estimated by both parsimony and maximum likelihood, were not significantly different \citep{hough2014}. This further suggests that the low diversity on the Y is not likely due to a lower mutation rate.

 A highly reduced level of Y chromosome variability is a predicted outcome of models Y-chromosome evolution based on the removal of deleterious mutations and linked neutral variants (e.g., background selection), ...and several cases.... but the occurrence of female-biased sex ratios in this species has also been predicted to lower Y diversity. Moreover, a further reduction in Y diversity could occure because of high variance in male reproductive success (Figure 1). This is not unusual in annual species such as \textit{R. hastatulus}s, which commonly exhibit extensive phenotypic plasticity in plant size and flower production (\X Harper 1977). Conceivably, such plastic differences among plants, owing to spatial heterogeneity in resources, could influence the intensity of male-male competition, and thus affect variance in male reproductive success. Given that male plants in this wind-pollinated species produce large amounts of pollen, and that female flowers are uniovulate, one might expect strong competition among males.

 In common with most flowering plants we do not have marker-based estimates of the variance in male reproductive success in \textit{R. hastatulus}. However, ... by modeling...we evaluated whether such an effect could result in a level of Y/A diversity similar to what we observed by comparing our results to predictions from a  model for the joint effects of sex ratio bias and male reproductive variance (see Methods). Estimates of sex ratio bias in \textit{R. hastatulus} have been estimated, and they are consistently female-biased, ranging from $N_{m}/(N_{m}+N_{f})=0.4$ to $N_{m}/(N_{m}+N_{f})=0.35$ (with a mean of 0.62; N=46 \citep{pickup2013influence}.The Y/A diversity ratio predicted from this sex ratio, assuming a Poisson distribution of offspring numbers, is approximately 0.2 (Table1, Figure 3). Our estimated mean Y/A ratio, however, is 0.02, which is significantly lower ($\textit{P}<0.0001$). Indeed, even for higher values for the variance in male reproductive success (\X), the corresponding predictions for Y/A are all significantly higher than our observed ratio (\X table). Finally, neutral models that could not be rejected, i.e., in which a highly female-biased sex ratio (\X) and a high variance in male reproductive success predicted an Y/A ratio that fit our data (\X), simultaneously predicts X/A ratios that are significantly different from what we observed (\X Figure). Thus, our results indicate that the combined effects of sex ratio bias and variance in reproductive success cannot jointly explain our observed levels of X and Y diversity in \textit{R. hastatulus}.

 %e.g., X/A diversity ratio that is 30% larger than what we found for the Texas race (X/ATX=0.85), and 80% larger than  the X/A ratio in the North Carolina race (X/ANC=0.35, further discussed below).

@@
